# NHANES: Initial Exploring {#dataviz} 

Next, we'll explore some data from the US National Health and Nutrition Examination Survey, or NHANES. 

We'll display R code as we go, but we'll return to all of the key coding ideas involved later in the Notes.

## The NHANES data: A First Sample

The `NHANES` package provides a sample of 10,000 NHANES responses from the 2009-10 and 2011-12 administrations, in a data frame also called `NHANES`. We can obtain the dimensions of this data frame (think of it as a rectangle of data) with the `dim()` function.

```{r}
dim(NHANES)
```

We see that we have `r nrow(NHANES)` rows and `r ncol(NHANES)` columns in the `NHANES` data frame.

For the moment, let's gather a random sample of 1,000 responses from the `r nrow(NHANES)` rows listed in the `NHANES` data frame, and then identify several variables of interest about those subjects^[For more on the NHANES data available in the NHANES package, type ?NHANES in the Console in R Studio.]. Some of the motivation for this example came from a Figure in @BaumerKaplanHorton.

```{r create_nh_dat1}
# library(NHANES) # already loaded NHANES package/library of functions, data

set.seed(431001) 
# use set.seed to ensure that we all get the same random sample 
# of 1,000 NHANES subjects in our nh_data collection

nh_dat1 <- 
    slice_sample(NHANES, n = 1000, replace = FALSE) %>%
    select(ID, SurveyYr, Gender, Age, Height)

nh_dat1
```

We have `r nrow(nh_dat1)` rows (observations) and `r ncol(nh_dat1)` columns (variables) that describe the responses listed in the rows. 

## Age and Height

Suppose we want to visualize the relationship of Height and Age in our 1,000 NHANES observations. The best choice is likely to be a scatterplot.

```{r nh_dat1_heightbyage1-fig}
ggplot(data = nh_dat1, aes(x = Age, y = Height)) +
    geom_point()
```

We note several interesting results here.

1. As a warning, R tells us that it has "Removed `r 1000 - sum(!is.na(nh_dat1$Age) & !is.na(nh_dat1$Height))` rows containing missing values (geom_point)." Only `r sum(!is.na(nh_dat1$Age) & !is.na(nh_dat1$Height))` subjects plotted here, because the remaining `r 1000 - sum(!is.na(nh_dat1$Age) & !is.na(nh_dat1$Height))` people have missing (NA) values for either Height, Age or both. 
2. Unsurprisingly, the measured Heights of subjects grow from Age `r min(nh_dat1$Age)` to Age 20 or so, and we see that a typical Height increases rapidly across these Ages. The middle of the distribution at later Ages is pretty consistent at at a Height somewhere between 150 and 175. The units aren't specified, but we expect they must be centimeters. The Ages are clearly reported in Years.
3. No Age is reported over 80, and it appears that there is a large cluster of Ages at 80. This may be due to a requirement that Ages 80 and above be reported at 80 so as to help mask the identity of those individuals.^[If you visit the NHANES help file with ?NHANES, you will see that subjects 80 years or older were indeed recorded as 80.]

As in this case, we're going to build most of our visualizations using tools from the `ggplot2` package, which is part of the `tidyverse` series of packages. You'll see similar coding structures throughout this Chapter, most of which are covered as well in Chapter 3 of @R4DS.


## Subset of Subjects with Known Age and Height

Before we move on, let's manipulate the data set a bit, to focus on only those subjects who have complete data on both Age and Height. This will help us avoid that warning message.

```{r create_nh_dat2_attempt1}
nh_dat2 <- nh_dat1 %>%
    filter(complete.cases(Age, Height)) 

summary(nh_dat2)
```

Note that the units and explanations for these variables are contained in the NHANES help file, available via typing `?NHANES` in the Console of R Studio, or by typing `NHANES` into the Search bar in R Studio's Help window.

## The Distinction between `Gender` and `Sex`

The `Gender` variable here is mis-named. These data refer to the biological status of these subjects, which is their `Sex`, and not the social construct of `Gender` which can be quite different. In our effort to avoid further confusion, we'll rename the variable `Gender` to `Sex` so as to more accurately describe what is actually measured here. 

To do this, we can use this approach...

```{r create_nhdat2_corrected}
nh_dat2 <- nh_dat1 %>%
    rename(Sex = Gender) %>%
    filter(complete.cases(Age, Height)) 

summary(nh_dat2)
```

That's better. How many observations do we have now? We could use `dim` to find out the number of rows and columns in this new data set.

```{r}
dim(nh_dat2)
```

Or, we could simply list the data set and read off the result.

```{r}
nh_dat2
```

## Age-Height and Sex?

Let's add Sex to the plot using color, and also adjust the y axis label to incorporate the units of measurement. 

```{r nh_dat2_heightbyageandsex1-fig}
ggplot(data = nh_dat2, aes(x = Age, y = Height, color = Sex)) +
    geom_point() +
    labs(title = "Height-Age Relationship in NHANES sample", 
         y = "Height in cm.")
```

### Can we show the Female and Male relationships in separate panels?

Sure.

```{r nh_dat2_heightbyageandsex2-fig}
ggplot(data = nh_dat2, aes(x = Age, y = Height, color = Sex)) +
    geom_point() + 
    labs(title = "Height-Age Relationship in NHANES sample", 
         y = "Height in cm.") +
    facet_wrap(~ Sex)
```

### Can we add a smooth curve to show the relationship in each plot?

Yep, and let's change the theme of the graph to remove the gray background, too.

```{r nh_dat2_heightbyageandsex3-fig}
ggplot(data = nh_dat2, aes(x = Age, y = Height, color = Sex)) +
    geom_point() + 
    geom_smooth(method = "loess", formula = y ~ x) +
    labs(title = "Height-Age Relationship in NHANES sample", 
         y = "Height in cm.") +
    theme_bw() +
    facet_wrap(~ Sex)
```


### What if we want to assume straight line relationships?

We could look at a linear model in the plot. Does this make sense here?

```{r nh_dat2_heightbyageandsex4-fig}
ggplot(data = nh_dat2, aes(x = Age, y = Height, color = Sex)) +
    geom_point() + 
    geom_smooth(method = "lm", formula = y ~ x) +
    labs(title = "Height-Age Relationship in NHANES sample", 
         y = "Height in cm.") +
    theme_bw() +
    facet_wrap(~ Sex)
```

I hope it seems clear from the graphs that a more complex relationship is going on here than just a straight line.

In the next Section of these Notes, we'll take a more carefully selected sample of NHANES respondents, and study those subjects in greater detail.